---
title: "Martin's key concepts"
subtitle: "The 20 concepts on the key-concepts list for GWAS, Mendelian randomization and proteomics, one module each"
engine: knitr
---

```{=html}
<script>document.body.classList.add("block-gwas");</script>
```

On 2 October the course published **Martin's key concepts**, a list of 20 concepts for the GWAS, Mendelian randomization (MR) and proteomics part of the course (Lectures 4–6, 8 and 9). When the lecture pages were written this list did not exist, so the material is spread across many sections and some concepts only got a sentence or two.

This page collects the list into **20 short modules**, in the list's own order. Each module gives an exam-ready definition you can say in one or two sentences, the common confusion to avoid, a quick check, and links to the lecture sections that go deeper. Three concepts that were thin in the lectures get more space here: **additive, dominant and recessive models** (K.4), **colocalization** (K.19) and **biomarker vs causal protein** (K.20).

Mark each module **Reviewed** or **Confident** as you go. The status is saved in this browser and counted on the [dashboard](dashboard.qmd) under 🔑.

::: {.callout-note collapse="true"}
## Source and what changed in the rest of the app
**Source:** "Martins part key_concepts.docx" (course material, uploaded 2 October), a list of 20 concept names without further text. The definitions, examples and questions on this page are written for this app.

**Changes elsewhere:** lecture sections that cover these concepts now carry the badge <span class="flag-badge flag-keyconcept">🔑 Martin's key concepts</span> instead of ❔ or 🗣️. Seminar topics (🗣️) are no longer described as "not on the written exam": seminar content can also come on the exam. The seminar's own questions are on [Seminar questions](seminar-questions.qmd).
:::

# Part 1: From SNP to GWAS

## K.1 SNP (single-nucleotide polymorphism) {#k-1 .topic flag="keyconcept"}

::: {.callout-tip icon="false"}
## Say it in one sentence
A **SNP** is a single base position in the genome where people in a population carry different nucleotides (alleles), for example some have C and others T; each SNP has an identifier such as an **rs number** from dbSNP.
:::

- Usually **biallelic**: two alleles, often called the **reference/other** and the **effect** allele in GWAS files.
- "Polymorphism" traditionally means common enough to be a normal variant in the population (often minor allele frequency ≥ 1%). A rare change in one family is usually called a **variant** or **mutation**.
- SNPs are what SNP arrays genotype and what GWAS tests, one at a time.

**Don't confuse:** a SNP is a *position* with variation in the population; an **allele** is one of the possible bases at that position; a **genotype** is the pair of alleles one person carries.

Deeper: [3.3c dbSNP and rs numbers](lecture3.qmd#t3-3c) · [4.2 SNPs, alleles, MAF](lecture4.qmd#t4-2)

:::: {.quiz}
Which statement describes a SNP correctly?

- A deletion of one gene that only some people carry
- **A single base position where alleles differ between people**
- A mutation that arises in a tumour but not in the germline
- The one allele that a person has at a given position

::: {.exp}
A SNP is a position with variation across the population. One person's pair of alleles at it is their genotype.
:::
::::

## K.2 Genotype and allele dosage (0/1/2) {#k-2 .topic flag="keyconcept"}

::: {.callout-tip icon="false"}
## Say it in one sentence
The **genotype** is the pair of alleles a person carries at a SNP; the **allele dosage** codes it as the number of copies of the effect allele, **0, 1 or 2**, which is the X variable in the GWAS regression.
:::

::: {.worked}
| Genotype (effect allele = T) | CC | CT | TT |
|---|---|---|---|
| Dosage | 0 | 1 | 2 |
:::

- Imputed SNPs give an **expected dosage** between 0 and 2 (for example 1.7), because the genotype is predicted with uncertainty.
- The dosage only means something once you know **which allele is counted**. Swap the effect allele and every dosage becomes 2 − dosage, and β changes sign.

**Don't confuse:** dosage counts the **effect allele**, which is not always the minor allele or the "risk" allele.

Deeper: [4.4 The additive model](lecture4.qmd#t4-4) · [A3 Reading association output](computer-exercises.qmd)

:::: {.quiz}
A file uses A as the effect allele. What is the dosage of a person with genotype GG?

- 2, because the person has two copies of an allele
- 1, because G is the other allele at this SNP
- **0, because the person has no copies of A**
- It cannot be coded without knowing the MAF

::: {.exp}
Dosage counts copies of the effect allele. GG has zero A alleles, so dosage = 0.
:::
::::

## K.3 Minor allele frequency (MAF) {#k-3 .topic flag="keyconcept"}

::: {.callout-tip icon="false"}
## Say it in one sentence
The **minor allele frequency** is the frequency of the less common allele at a SNP in a given population, so it is always **between 0 and 0.5**.
:::

- Computed from genotype counts: each heterozygote contributes one copy, each homozygote two. With $n_{AA}, n_{Aa}, n_{aa}$: $f_a = (2n_{aa} + n_{Aa}) / (2N)$.
- **Low MAF means low power**: few people carry the allele, so β is estimated with a large SE. Rare variants are why GWAS QC often filters on MAF.
- MAF is **population-specific**. The minor allele in Europeans can be the major allele in another ancestry, which matters when combining datasets (seminar question on GLGC vs 1000 Genomes).

**Don't confuse:** MAF is not the same as the **effect allele frequency (EAF)**. If the effect allele is the major allele, EAF = 1 − MAF.

Deeper: [4.2 SNPs, alleles, MAF and genotype frequencies](lecture4.qmd#t4-2) · [3.9 Hardy–Weinberg](lecture3.qmd#t3-9)

:::: {.quiz}
In 1,000 people there are 640 CC, 320 CT and 40 TT. What is the MAF?

- 0.04, the share of people who are TT
- 0.36, the share of people who carry a T
- **0.20, the share of all alleles that are T**
- 0.80, the share of all alleles that are C

::: {.exp}
T alleles = 2 × 40 + 320 = 400 out of 2,000 alleles = 0.20. T is the less common allele, so MAF = 0.20.
:::
::::

## K.4 Additive, dominant and recessive models {#k-4 .topic flag="keyconcept"}

::: {.callout-tip icon="false"}
## Say it in one sentence
The genetic model says **how the genotype is coded** in the regression: **additive** counts copies (0/1/2, each copy adds the same effect), **dominant** compares carriers with non-carriers (0/1/1), and **recessive** compares two copies with the rest (0/0/1).
:::

::: {.worked}
**Coding of the same SNP under the three models** (effect allele = T)

| Model | CC | CT | TT | What β means |
|---|---|---|---|---|
| **Additive** | 0 | 1 | 2 | Change in phenotype **per extra copy** of T |
| **Dominant** (T dominant) | 0 | 1 | 1 | Difference between **carriers** (one or two T) and non-carriers |
| **Recessive** (T recessive) | 0 | 0 | 1 | Difference between **TT** and everyone else |
:::

```{r}
#| echo: false
#| fig-width: 7.5
#| fig-height: 2.8
#| fig-cap: "Mean phenotype by genotype under each model, with the same effect of 1 unit. Additive: a straight line through 0, 1, 2. Dominant: the heterozygote already looks like TT. Recessive: only TT differs."
library(ggplot2)
d <- data.frame(
  model = factor(rep(c("Additive", "Dominant", "Recessive"), each = 3),
                 levels = c("Additive", "Dominant", "Recessive")),
  genotype = factor(rep(c("CC", "CT", "TT"), 3), levels = c("CC", "CT", "TT")),
  mean = c(0, 1, 2,  0, 1, 1,  0, 0, 1)
)
ggplot(d, aes(genotype, mean, group = 1)) +
  geom_line(colour = "#2f5d8a", linewidth = 1) +
  geom_point(size = 3, colour = "#2f5d8a") +
  facet_wrap(~ model) +
  labs(x = NULL, y = "Mean phenotype") +
  theme_minimal(base_size = 12)
```

**Why GWAS use the additive model** (Lecture 4, "Worth noting 2"): you test a million SNPs and do not know each one's true pattern. The additive model is the simple default; it has reasonable power even when the true pattern is dominant or recessive, and it uses one degree of freedom. Dominant and recessive models are useful when you study **one specific SNP or gene** in depth.

**Don't confuse:** dominance describes the **genotype–phenotype relationship**, not the allele. The same allele can act dominant for one phenotype and recessive for another (Lecture 4).

Deeper: [4.4 The additive model](lecture4.qmd#t4-4) · [3.6 Mendelian genetics](lecture3.qmd#t3-6)

:::: {.quiz}
Under a recessive model for allele T, which people are compared?

- Each extra copy of T is compared with one fewer copy
- People with at least one T against people with no T
- **People with two copies of T against everyone else**
- Heterozygotes against both kinds of homozygote

::: {.exp}
Recessive coding is 0/0/1: only TT is coded 1, so β is the difference between TT and the CC + CT group.
:::
::::

:::: {.quiz}
The mean phenotype is 4.0 in CC, 5.0 in CT and 5.0 in TT. Which model fits this pattern best?

- **Dominant, since one copy has the full effect**
- Additive, since each copy adds one unit
- Recessive, since only TT differs from CC
- None, since the heterozygote should lie halfway

::: {.exp}
CT already has the same mean as TT, so one copy is enough: a dominant pattern (0/1/1). An additive pattern would give about 4.0, 4.5, 5.0 or 4, 5, 6.
:::
::::

## K.5 Genome-wide association study (GWAS) {#k-5 .topic flag="keyconcept"}

::: {.callout-tip icon="false"}
## Say it in one sentence
A **GWAS** tests each of roughly a million SNPs across the genome, one at a time, for association with a phenotype, using a regression of the phenotype on allele dosage adjusted for covariates such as age, sex and genetic principal components.
:::

- **Hypothesis-free**: no candidate gene is chosen in advance.
- Output: one β, SE and p per SNP (summary statistics, K.11), shown as a **Manhattan plot** (position vs −log₁₀ p) and checked with a **QQ plot**.
- Continuous phenotype → linear regression (β = mean difference per allele); binary → logistic regression (β = log-odds per allele, OR = e^β).
- A GWAS finds **associations**. The top SNP is usually a **tag** in LD with the causal variant, and the association can still be confounded (population structure).

**Don't confuse:** GWAS finds **loci**, not genes or mechanisms. Going from a hit to biology needs fine-mapping, QTL data, colocalization, MR or experiments.

Deeper: [4.6 The GWAS scan](lecture4.qmd#t4-6) · [4.9 Population structure](lecture4.qmd#t4-9) · [4.10 From hit to biology](lecture4.qmd#t4-10) · [5.2 The full GWAS pipeline](lecture5.qmd#t5-2)

:::: {.quiz}
Why are genetic principal components included as covariates in a GWAS?

- To increase the number of SNPs that can be tested
- **To adjust for ancestry differences that can confound**
- To convert the additive model into a dominant one
- To remove SNPs that are in strong LD with each other

::: {.exp}
If allele frequencies and the phenotype both differ between ancestry groups, SNPs look associated because of ancestry. The PCs capture ancestry and remove that confounding.
:::
::::

## K.6 Effect size, standard error and p-value {#k-6 .topic flag="keyconcept"}

::: {.callout-tip icon="false"}
## Say it in one sentence
The **effect size β** is the estimated change in the phenotype per copy of the effect allele (log-odds for a binary outcome), the **standard error** says how precisely β is estimated, and the **p-value** measures the evidence against β = 0, based on $z = \beta / SE$.
:::

::: {.worked}
**Example.** β = 0.05 SD per allele, SE = 0.008.

| Step | Value |
|---|---|
| z = β / SE | 0.05 / 0.008 = 6.25 |
| Two-sided p | ≈ 4 × 10⁻¹⁰, below 5 × 10⁻⁸ |
| 95% CI | 0.05 ± 1.96 × 0.008 = 0.034 to 0.066 |
:::

- **SE shrinks** with larger sample size, higher MAF and lower phenotype variance. That is why a tiny β can have a tiny p in a large GWAS.
- **Binary outcome:** β is a log(OR). OR per allele = e^β; for two copies under the additive model, OR = e^(2β).
- **p measures evidence, not importance.** A highly significant SNP can explain almost none of the phenotype variance.

**Don't confuse:** a small p-value does not mean a large effect. Compare β (and its units) to judge size.

Deeper: [4.3 Phenotype types and the per-allele effect](lecture4.qmd#t4-3) · [4.4 β, SE and p](lecture4.qmd#t4-4)

<div class="drill" data-drill="or"></div>

:::: {.quiz}
Two SNPs have the same β, but SNP 2 has a much smaller SE. What is the most likely reason?

- SNP 2 has a larger effect on the phenotype
- **SNP 2 has a higher MAF or a larger sample**
- SNP 2 lies closer to the gene it regulates
- SNP 2 was tested with the dominant model

::: {.exp}
SE depends on sample size, allele frequency and phenotype variance. With β the same, a smaller SE comes from more information: more people or more carriers of the allele.
:::
::::

## K.7 Genome-wide significance {#k-7 .topic flag="keyconcept"}

::: {.callout-tip icon="false"}
## Say it in one sentence
**Genome-wide significance** is the conventional threshold **p < 5 × 10⁻⁸** for a GWAS hit; it is a Bonferroni correction of α = 0.05 for about **one million independent common-variant tests** in the genome.
:::

$$\frac{0.05}{1\,000\,000} = 5 \times 10^{-8}$$

- "Independent" is the key word: there are many more SNPs than that, but because of **LD** neighbouring SNPs are correlated, so the effective number of independent tests in European-ancestry genomes is about one million.
- It controls the **family-wise error rate**: the chance of even one false positive across the whole scan is about 5%.
- SNPs with 5 × 10⁻⁸ < p < 10⁻⁵ are sometimes called "suggestive".

**Don't confuse:** passing the threshold does not make a SNP causal, and failing it does not mean "no effect"; it may mean too little power.

Deeper: [4.7 Multiple testing, Manhattan and QQ plots](lecture4.qmd#t4-7) · [7.11 FWER vs FDR](lecture7.qmd#t7-11) · [7.12 Bonferroni](lecture7.qmd#t7-12)

:::: {.quiz}
Where does the threshold 5 × 10⁻⁸ come from?

- It is the p-value of the strongest SNP in the first GWAS
- **0.05 divided by about a million independent tests**
- 0.05 divided by the total number of SNPs on an array
- It is the FDR level that Benjamini–Hochberg gives

::: {.exp}
Bonferroni: α / m with m ≈ 10⁶ independent tests, after accounting for LD. It is a FWER threshold, not an FDR threshold.
:::
::::

## K.8 Linkage disequilibrium (LD; r²) {#k-8 .topic flag="keyconcept"}

::: {.callout-tip icon="false"}
## Say it in one sentence
**Linkage disequilibrium** is the non-random association between alleles at nearby SNPs in a population, so knowing one SNP's allele tells you something about the other's; **r²** is the squared correlation between the two SNPs' allele dosages, from 0 (independent) to 1 (perfect proxy).
:::

- LD exists because nearby SNPs are **inherited together**: recombination between them is rare, so the haplotypes they sat on are preserved over generations.
- **r² = 1**: one SNP perfectly predicts the other, so they give the same GWAS signal. **r² ≈ 0**: independent.
- LD is **population-specific**. Compute it in a reference panel that matches the ancestry of the GWAS (for example the European part of 1000 Genomes).
- LD generally decays with distance, but **physical distance is not LD**: two close SNPs can have low r², and LD blocks vary in size.

**Don't confuse:** **linkage** is about co-inheritance within families (recombination frequency); **LD** is the correlation of alleles across a population.

Deeper: [3.7 Linkage, recombination and LD](lecture3.qmd#t3-7) · [4.8 Lead SNP and r²](lecture4.qmd#t4-8) · [5.4b r² bands](lecture5.qmd#t5-4b)

<div class="drill" data-drill="ld"></div>

:::: {.quiz}
SNP A is genome-wide significant. SNP B, 20 kb away, has r² = 0.95 with A. What do you expect for SNP B?

- **A similar association, since B almost copies A**
- No association, because only one SNP can be causal
- An association in the opposite direction from A
- An association only if B lies inside a gene's exon

::: {.exp}
With r² = 0.95, B's alleles nearly predict A's, so B shows almost the same signal. That is why GWAS hits appear as towers of correlated SNPs.
:::
::::

## K.9 Lead SNP and GWAS locus {#k-9 .topic flag="keyconcept"}

::: {.callout-tip icon="false"}
## Say it in one sentence
A **GWAS locus** is a genomic region with a significant association signal, usually many correlated SNPs; the **lead SNP** is the SNP with the **smallest p-value** in that locus and is used to represent it.
:::

- A locus is defined around the lead SNP, by a distance window (for example ± 500 kb) and/or by LD with the lead SNP.
- The lead SNP is the **best statistical tag**, not necessarily the causal variant. The causal variant can be any SNP in strong LD with it, or one that was not genotyped at all.
- One locus can contain **more than one independent signal** (conditional analysis or clumping can reveal a secondary signal).

**Don't confuse:** "lead SNP" (strongest p in the region) ≠ "causal SNP" (the variant that changes the biology) ≠ "nearest gene" (often assumed, not proven).

Deeper: [4.8 Lead SNP and r²](lecture4.qmd#t4-8) · [4.10 Locus, lead SNP, QTLs](lecture4.qmd#t4-10) · [5.4 Finding the causal SNP](lecture5.qmd#t5-4)

:::: {.quiz}
Why is the lead SNP not automatically the causal variant?

- Because the lead SNP is always in a non-coding region
- **Because SNPs in strong LD with it give nearly the same p**
- Because the lead SNP is chosen by its position in the gene
- Because causal variants never reach genome-wide significance

::: {.exp}
Many SNPs in high LD carry almost the same signal, and the smallest p among them can be partly chance or better genotyping. The causal variant may be another SNP in LD or an unmeasured one.
:::
::::

## K.10 LD clumping and independent genetic signals {#k-10 .topic flag="keyconcept"}

::: {.callout-tip icon="false"}
## Say it in one sentence
**LD clumping** reduces a list of significant SNPs to **independent signals**: take the most significant SNP, remove all SNPs within a window that are in LD with it above an r² threshold, and repeat with the next most significant remaining SNP.
:::

::: {.worked}
**The algorithm**

| Step | Action |
|---|---|
| 1 | Sort significant SNPs by p, smallest first |
| 2 | Take the top SNP as an **index SNP** |
| 3 | Remove SNPs within the window (for example 1,000 kb) with r² above the threshold (for example 0.01–0.1) |
| 4 | Repeat with the next remaining SNP until none are left |
:::

- **Why:** correlated SNPs carry the same information. For MR you need **independent** instruments, otherwise the same signal is counted several times and the SE is too small. For PRS clumping avoids double-counting.
- **Choices matter:** a **wider window** compares more SNP pairs (slower, catches long-range LD, leaves fewer SNPs); a **lower r² threshold** removes more SNPs (stricter independence).
- More SNPs can survive clumping than there are causal variants: two weakly correlated tags of one causal variant can both pass a lenient threshold, and SNPs outside the window are never compared.

**Don't confuse:** "independent" (low r² between the kept SNPs) is not "causal". Each clump still represents a region.

Deeper: [5.4b r² bands and LD clumping](lecture5.qmd#t5-4b) · [A5 Manual LD clumping](computer-exercises.qmd) · [Seminar S.1 and S.5](seminar-questions.qmd)

:::: {.quiz}
Why must MR instruments be clumped before the IVW estimate?

- Clumping makes every instrument genome-wide significant
- **Correlated SNPs would count one signal several times**
- Clumping removes the need to harmonise the alleles
- IVW can only handle one SNP per chromosome at most

::: {.exp}
IVW treats each SNP as independent evidence. SNPs in LD repeat the same signal, so the estimate gets falsely precise.
:::
::::

## K.11 GWAS summary statistics {#k-11 .topic flag="keyconcept"}

::: {.callout-tip icon="false"}
## Say it in one sentence
**GWAS summary statistics** are the per-SNP results of a GWAS (SNP ID, chromosome, position, effect and other allele, allele frequency, β, SE, p and N) with **no individual-level data**, which is why they can be shared publicly and reused, for example in two-sample MR or PRS.
:::

- Before using them, check four things: **which allele is the effect allele**, whether β is **OR or log(OR)**, the **scale** of the phenotype (SD units? log?), and the **genome build** (hg19/GRCh37 vs hg38/GRCh38).
- **LD cannot be computed** from summary statistics: they have no genotypes per person. Use a reference panel with matching ancestry.
- Two-sample MR needs two sets: one for the exposure, one for the outcome, matched on SNP and allele.

**Don't confuse:** summary statistics are per SNP, not per person. You cannot rerun the regression with new covariates from them.

Deeper: [4.6 Summary statistics](lecture4.qmd#t4-6) · [5.5 LD with summary statistics](lecture5.qmd#t5-5) · [9.2 The data you need](lecture9.qmd#t9-2)

:::: {.quiz}
You have summary statistics only. Which task needs an external reference panel?

- Reading off the β and p of a given SNP
- Finding SNPs below p = 5 × 10⁻⁸
- **Estimating r² between two SNPs for clumping**
- Converting a log-odds β into an odds ratio

::: {.exp}
r² needs the genotypes of individuals. Summary statistics have none, so LD comes from a reference panel such as 1000 Genomes.
:::
::::

# Part 2: Mendelian randomization

## K.12 Instrumental variable (IV) {#k-12 .topic flag="keyconcept"}

::: {.callout-tip icon="false"}
## Say it in one sentence
An **instrumental variable** is a variable that shifts the exposure, shares no causes with the outcome, and affects the outcome only through the exposure; it lets you estimate a causal effect even when the exposure–outcome relationship is confounded.
:::

```
   Z (instrument) ───▶ X (exposure) ───▶ Y (outcome)
                          ▲                ▲
                          └──── U (confounders) ┘
```

- The idea comes from econometrics. The classic example is the **randomization** in an RCT: the coin flip shifts treatment and is independent of everything else.
- In MR the instrument is a **genetic variant**: alleles are allocated at conception, before the exposure and the disease, and independently of most lifestyle confounders.

**Don't confuse:** an IV is not a confounder or a covariate to adjust for. Its job is to give variation in the exposure that confounders do not touch.

Deeper: [6.3 Genetic variants as instruments](lecture6.qmd#t6-3) · [6.4 Why we need MR: confounding](lecture6.qmd#t6-4)

:::: {.quiz}
In a randomized trial, what plays the role of the instrumental variable?

- The outcome measured at the end of follow-up
- The confounders recorded at the baseline visit
- **The random allocation to treatment or control**
- The treatment the participant actually took

::: {.exp}
Allocation shifts the treatment received and is independent of confounders, just like an allele in MR. The treatment actually taken is the exposure.
:::
::::

## K.13 Mendelian randomization (MR) {#k-13 .topic flag="keyconcept"}

::: {.callout-tip icon="false"}
## Say it in one sentence
**Mendelian randomization** uses genetic variants associated with an exposure as instrumental variables to estimate the **causal effect** of the exposure on an outcome, exploiting that alleles are randomly allocated at conception, which limits confounding and rules out reverse causation.
:::

- "Nature's randomized trial": people who inherited exposure-raising alleles are, on average, like people who were randomized to a slightly higher exposure for life.
- **One-sample MR**: exposure and outcome in the same people. **Two-sample MR**: SNP–exposure and SNP–outcome effects from two different GWAS, using only summary statistics.
- The answer is the effect per unit of **genetically predicted** exposure, which is lifelong and usually small; it is not the same as the effect of a drug given in middle age.

**Don't confuse:** MR is only as good as its three assumptions (K.14). A significant MR estimate with a pleiotropic instrument is not causal evidence.

Deeper: [6.4 Why we need MR](lecture6.qmd#t6-4) · [9.1 Two-sample MR](lecture9.qmd#t9-1) · [9.7 Interpreting an MR estimate](lecture9.qmd#t9-7)

:::: {.quiz}
Why does MR rule out reverse causation?

- Because the outcome GWAS is always run before the exposure GWAS
- Because MR uses only SNPs located inside the exposure's gene
- **Because the disease cannot change the inherited genotype**
- Because the Wald ratio is always positive for causal effects

::: {.exp}
The genotype is fixed at conception. Disease can change measured biomarkers but not the alleles, so the arrow cannot run from outcome to instrument.
:::
::::

## K.14 The three MR assumptions {#k-14 .topic flag="keyconcept"}

::: {.callout-tip icon="false"}
## Say it in one sentence
A valid MR instrument must be (1) **relevant**: robustly associated with the exposure; (2) **independent**: not associated with confounders of the exposure–outcome relationship; and (3) satisfy the **exclusion restriction**: affect the outcome only through the exposure.
:::

::: {.worked}
| # | Assumption | How you check it | Typical threat |
|---|---|---|---|
| 1 | **Relevance** | p < 5 × 10⁻⁸ in the exposure GWAS; **F-statistic > 10** | Weak instruments |
| 2 | **Independence** | Cannot be proven; adjust for ancestry, check associations with known confounders | Population stratification, assortative mating, selection |
| 3 | **Exclusion restriction** | Cannot be proven; biology of the gene, MR-Egger intercept, weighted median, outlier tests | **Horizontal pleiotropy** |
:::

- Only assumption 1 can be tested directly. Assumptions 2 and 3 are made **plausible**, never proven, which is why MR uses several sensitivity analyses.

**Don't confuse:** "the SNP is associated with the outcome" is *not* an assumption. That is what MR tests, and it is only interesting if the three assumptions hold.

Deeper: [6.5 The three MR assumptions](lecture6.qmd#t6-5) · [6.8 Sensitivity analyses](lecture6.qmd#t6-8) · [9.3 F-statistic](lecture9.qmd#t9-3)

:::: {.quiz}
Which MR assumption can be checked directly in the data?

- **Relevance, from the SNP–exposure association**
- Independence, from the SNP–outcome association
- Exclusion restriction, from the MR-Egger slope
- None of them, since all three are untestable

::: {.exp}
Relevance is visible in the exposure GWAS (p-value, F-statistic). Independence and exclusion restriction involve unmeasured paths and can only be made plausible.
:::
::::

## K.15 Horizontal vs vertical pleiotropy {#k-15 .topic flag="keyconcept"}

::: {.callout-tip icon="false"}
## Say it in one sentence
**Pleiotropy** is one variant affecting several traits. It is **vertical** when the variant affects the outcome *through* the exposure (fine for MR), and **horizontal** when it affects the outcome through a separate pathway that bypasses the exposure (violates the exclusion restriction).
:::

```
Vertical (OK):     SNP ──▶ LDL-C ──▶ atherosclerosis ──▶ CHD

Horizontal (bad):  SNP ──▶ LDL-C ──────────────────────▶ CHD
                     └──▶ inflammation ────────────────▶ CHD
```

- Horizontal pleiotropy shows up as instruments whose Wald ratios disagree with the rest, an MR-Egger intercept away from zero, or IVW and weighted median giving different answers.
- **trans-pQTLs** and SNPs in genes with many functions are the usual suspects. **cis** instruments in the exposure's own gene reduce the risk.

**Don't confuse:** vertical pleiotropy is not a problem to remove. It is the causal chain MR is designed to detect.

Deeper: [6.6 Pleiotropy](lecture6.qmd#t6-6) · [6.8 MR-Egger and weighted median](lecture6.qmd#t6-8) · [9.8 Pleiotropy and pitfalls](lecture9.qmd#t9-8)

:::: {.quiz}
A blood pressure SNP also raises BMI, and BMI raises stroke risk directly. In an MR of blood pressure on stroke this is:

- Vertical pleiotropy, which does not bias the estimate
- **Horizontal pleiotropy, which can bias the estimate**
- A weak-instrument problem under the relevance assumption
- Reverse causation from stroke to blood pressure

::: {.exp}
The SNP reaches stroke through BMI, a path that does not go through blood pressure. That is horizontal pleiotropy and breaks the exclusion restriction.
:::
::::

## K.16 Wald ratio and inverse-variance weighted (IVW) MR {#k-16 .topic flag="keyconcept"}

::: {.callout-tip icon="false"}
## Say it in one sentence
With one instrument the causal estimate is the **Wald ratio**, the SNP's effect on the outcome divided by its effect on the exposure; with several independent instruments the **IVW estimate** is a precision-weighted average of the Wald ratios, equal to a weighted regression of β-outcome on β-exposure **through the origin**.
:::

$$\hat\beta_{\text{Wald}} = \frac{\beta_{\text{SNP}\to Y}}{\beta_{\text{SNP}\to X}}, \qquad SE_{\text{Wald}} \approx \frac{SE_{\text{SNP}\to Y}}{|\beta_{\text{SNP}\to X}|}$$

$$\hat\beta_{\text{IVW}} = \frac{\sum_j \beta_{X_j}\beta_{Y_j}/SE_{Y_j}^2}{\sum_j \beta_{X_j}^2/SE_{Y_j}^2}$$

::: {.worked}
**Example.** Three clumped SNPs (exposure in mmol/L, outcome log-odds).

| SNP | β exposure | β outcome | SE outcome | Wald ratio |
|---|---|---|---|---|
| 1 | 0.20 | 0.10 | 0.02 | 0.50 |
| 2 | 0.10 | 0.06 | 0.02 | 0.60 |
| 3 | 0.30 | 0.12 | 0.03 | 0.40 |

IVW numerator = (0.20·0.10 + 0.10·0.06)/0.0004 + 0.30·0.12/0.0009 = 65 + 40 = 105. Denominator = (0.04 + 0.01)/0.0004 + 0.09/0.0009 = 125 + 100 = 225. IVW = 105/225 ≈ **0.47** log-odds per mmol/L, OR ≈ e^0.47 ≈ 1.6.
:::

- Before computing: **harmonise** the alleles so both β refer to the same effect allele, and for a binary outcome use **log(OR)**.
- Through the origin = assumes **no horizontal pleiotropy** (zero exposure effect → zero outcome effect). MR-Egger frees the intercept to test that.

**Don't confuse:** the Wald ratio is per unit of the exposure's scale. An exposure in SD units and one in mmol/L give different numbers for the same biology.

Deeper: [6.7 The Wald ratio and IVW](lecture6.qmd#t6-7) · [9.5 Harmonisation](lecture9.qmd#t9-5) · [9.6 Wald ratio and IVW](lecture9.qmd#t9-6)

<div class="drill" data-drill="wald"></div>

<div class="drill" data-drill="ivw"></div>

:::: {.quiz}
Why does the IVW regression line go through the origin?

- Because the exposure GWAS centres all β values at zero
- **Because a SNP with no exposure effect should have no outcome effect**
- Because weighting by precision removes the intercept term
- Because it makes the estimate robust to pleiotropic SNPs

::: {.exp}
Under the exclusion restriction the only way a SNP affects the outcome is through the exposure, so zero on X must give zero on Y. A non-zero intercept (MR-Egger) suggests directional pleiotropy.
:::
::::

# Part 3: Proteomics and genetics

## K.17 Proteomics and protein abundance {#k-17 .topic flag="keyconcept"}

::: {.callout-tip icon="false"}
## Say it in one sentence
**Proteomics** is the large-scale measurement of the proteins in a sample (the proteome); most platforms measure **protein abundance**, the amount or concentration of each protein, often as a relative value such as Olink's log₂-scale NPX, not its activity.
:::

- **Why proteins:** they do the work in cells, change with disease, diet, drugs and age, and are the targets of most drugs.
- **Why it is hard:** plasma protein concentrations span about **10 orders of magnitude**, so a few abundant proteins (albumin, immunoglobulins) dominate; one gene can give several **proteoforms** (isoforms, post-translational modifications).
- **Platforms:** affinity-based (Olink, SomaScan: antibodies or aptamers, thousands of proteins, relative units) and mass spectrometry (peptides, can see isoforms and give absolute amounts with internal standards).
- **NPX** is relative and log₂: +1 NPX = twice as much of that protein in that sample. It cannot be compared between different proteins.

**Don't confuse:** abundance ≠ function. A variant can change what a protein does without changing how much there is, and an affinity assay can miss a form it does not bind.

Deeper: [8.1 Why study the proteome](lecture8.qmd#t8-1) · [8.4 Three technology families](lecture8.qmd#t8-4) · [8.5 Units and NPX](lecture8.qmd#t8-5) · [8.13 Reading: comparing eight platforms](lecture8.qmd#t8-13)

:::: {.quiz}
Protein X has NPX 6 in sample A and NPX 8 in sample B. What can you say?

- **Sample B has about four times as much protein X**
- Sample B has 2 units/mL more protein X than sample A
- Sample B has more protein X than any protein in sample A
- Sample B's protein X is more biologically active

::: {.exp}
NPX is log₂, so a difference of 2 = 2² = 4-fold for the same protein. It says nothing about other proteins or about activity.
:::
::::

## K.18 pQTLs: cis-pQTL and trans-pQTL {#k-18 .topic flag="keyconcept"}

::: {.callout-tip icon="false"}
## Say it in one sentence
A **pQTL** (protein quantitative trait locus) is a genetic variant associated with the abundance of a protein; it is **cis** when it lies near the gene that encodes the protein (commonly within 500 kb–1 Mb) and **trans** when it lies far away or on another chromosome.
:::

::: {.worked}
| | **cis-pQTL** | **trans-pQTL** |
|---|---|---|
| Location | Near the protein's own gene | Far away (> 500 kb) or another chromosome |
| Likely mechanism | Direct: transcription, splicing, stability, secretion of that protein | Indirect: transcription factors, receptors, pathways |
| As MR instrument | **Preferred**: clear biology, less pleiotropy | Higher risk of horizontal pleiotropy |
:::

- Found with a GWAS that has the protein level as the phenotype: protein = β₀ + β₁ × dosage + ε.
- **Caution:** a coding (missense) cis variant can change the **epitope** an antibody or aptamer binds and fake a pQTL: lower measured signal, unchanged amount.

**Don't confuse:** cis is defined relative to the **protein's gene**, not relative to any gene. A SNP next to gene A that changes protein B is trans for B.

Deeper: [8.9 pQTLs](lecture8.qmd#t8-9) · [8.10 Protein MR](lecture8.qmd#t8-10) · [8.11 Challenges of protein MR](lecture8.qmd#t8-11)

:::: {.quiz}
A SNP 50 kb from gene *ABC* is associated with the level of protein ABC. What is it, and why does it matter?

- **A cis-pQTL, a good candidate MR instrument**
- A trans-pQTL, with a high risk of pleiotropy
- A cis-pQTL, which proves ABC causes disease
- An eQTL, since it is near a gene's promoter

::: {.exp}
Near the protein's own gene → cis. cis-pQTLs are preferred instruments for protein MR, but the three MR assumptions and colocalization still need checking.
:::
::::

## K.19 Colocalization {#k-19 .topic flag="keyconcept"}

::: {.callout-tip icon="false"}
## Say it in one sentence
**Colocalization** asks whether two association signals in the same region, for example a pQTL for a protein and a GWAS hit for a disease, are driven by the **same causal variant**, rather than by two different variants that happen to be close together and in LD.
:::

**The logic from the slides** (Lecture 8): compare the **pattern** of association across the region for the protein and for the disease. If one shared variant drives both, SNPs in LD with it produce the **same shape** of signal for both traits: the same SNPs are strongest, and the association weakens in step as LD with the shared variant falls.

```{r}
#| echo: false
#| fig-width: 7.5
#| fig-height: 3.2
#| fig-cap: "Simulated regional association plots. Left: one shared causal variant, so the protein and disease peaks have the same shape and the same top SNP. Right: two different causal variants 60 kb apart; the peaks overlap in the region but do not line up."
set.seed(19)
pos <- seq(0, 400, by = 4)
peak <- function(centre, height, width = 25) height * exp(-((pos - centre) / width)^2) + abs(rnorm(length(pos), 0, 0.5))
d <- rbind(
  data.frame(panel = "Shared variant (colocalize)", trait = "Protein (pQTL)", pos = pos, logp = peak(200, 30)),
  data.frame(panel = "Shared variant (colocalize)", trait = "Disease (GWAS)", pos = pos, logp = peak(200, 12)),
  data.frame(panel = "Two variants (do not colocalize)", trait = "Protein (pQTL)", pos = pos, logp = peak(170, 30)),
  data.frame(panel = "Two variants (do not colocalize)", trait = "Disease (GWAS)", pos = pos, logp = peak(230, 12))
)
ggplot(d, aes(pos, logp, colour = trait)) +
  geom_point(size = 1.4, alpha = 0.8) +
  facet_wrap(~ panel) +
  scale_colour_manual(values = c("#d9480f", "#2f5d8a"), name = NULL) +
  labs(x = "Position in region (kb)", y = expression(-log[10](p))) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

**Why MR alone is not enough:** if SNP A raises the protein and a different SNP B in LD with A raises disease risk through another mechanism, an MR using A will still find a "protein → disease" effect. That is **LD confounding**. Colocalization checks for it.

::: {.callout-note collapse="true"}
## Beyond the slides: the formal coloc method
The widely used Bayesian method (the R package *coloc*) compares five hypotheses for a region: **H0** no association with either trait, **H1** only trait 1, **H2** only trait 2, **H3** both traits but **different** causal variants, **H4** both traits and **one shared** causal variant. It reports a posterior probability for each, and **PP.H4 above about 0.8** is often taken as evidence of colocalization. It needs only summary statistics for both traits in the region. The slides describe the idea; this formal version is background, not course material.
:::

**Don't confuse:** "both signals are in the same locus" is not colocalization. Two causal variants 50 kb apart can sit in one locus with overlapping towers of association.

Deeper: [8.11 Challenges of protein MR (colocalization)](lecture8.qmd#t8-11) · [8.12 From association to mechanism](lecture8.qmd#t8-12)

:::: {.quiz}
A pQTL for protein P and a GWAS hit for disease D are in the same region, but the strongest pQTL SNPs are weak for D and vice versa. What is the most likely conclusion?

- The protein causes the disease, since the signals overlap
- **Two different causal variants drive the two signals**
- The pQTL must be a trans-pQTL for the protein P
- The disease GWAS must be confounded by ancestry

::: {.exp}
If one variant drove both, the same SNPs would be strongest for both traits. Mismatched peaks point to distinct causal variants, so an MR through P here may reflect LD, not causation.
:::
::::

## K.20 Biomarker vs causal protein {#k-20 .topic flag="keyconcept"}

::: {.callout-tip icon="false"}
## Say it in one sentence
A **biomarker** is a measurable molecule that **indicates** a state or risk (it can be a cause, a consequence or just a correlate of disease), whereas a **causal protein** is one where **changing its level or activity changes disease risk**, which is what you need for a drug target.
:::

::: {.worked}
| | **Biomarker** | **Causal protein** |
|---|---|---|
| Question | Does it **predict** or **reflect** disease? | Does changing it **change** disease? |
| Evidence | Association, prediction (AUC), reproducibility | MR with cis-pQTLs, colocalization, experiments, trials |
| Useful for | Diagnosis, prognosis, risk prediction, monitoring treatment | Drug targets, understanding mechanism |
| Example type | Troponin after heart damage: excellent marker, consequence not cause | A protein whose cis-pQTL predicts disease in MR and colocalizes |
:::

- A protein associated with disease in a PWAS can be (1) a **cause**, (2) a **consequence** of disease (reverse causation), or (3) a **correlate** sharing a confounder (smoking, BMI, inflammation).
- A good biomarker does **not** have to be causal; a causal protein is **not** always a good biomarker (it may vary little between people or be hard to measure).
- The route from biomarker to causal protein (Lecture 8): association → pQTL → MR and colocalization → experiments (CRISPR, cell models, animal models).

**Don't confuse:** "strongly associated" and "predictive" do not mean "causal". Lowering a consequence-marker does not undo the disease.

Deeper: [1.3b Biomarkers and their five roles](lecture1.qmd#t1-3b) · [8.7 PWAS](lecture8.qmd#t8-7) · [8.8 Association is still not causation](lecture8.qmd#t8-8) · [8.12 From association to mechanism](lecture8.qmd#t8-12)

:::: {.quiz}
A plasma protein rises sharply after a heart attack and predicts death well. What does this tell you?

- It causes heart attacks and should be a drug target
- **It is a useful biomarker, but may not be causal**
- It must have a strong cis-pQTL in the general population
- It proves that MR would find a causal effect on CHD

::: {.exp}
Rising after the event suggests it is a consequence (like troponin). Prediction makes it a useful biomarker; causality needs MR, colocalization or experiments.
:::
::::

# Check yourself: all 20 in one go

::: {.callout-tip collapse="true" icon="false"}
## Explain the chain from SNP to causal protein in eight sentences, using as many key concepts as you can.
**Model answer.** A **SNP** is a position where people carry different alleles; each person's **genotype** is coded as an **allele dosage** of 0, 1 or 2 copies of the effect allele, and the **MAF** says how common the rarer allele is. A **GWAS** regresses the phenotype on dosage for each SNP, usually under the **additive model** (dominant and recessive models are used for single SNPs), giving an **effect size, SE and p-value** per SNP. Hits below the **genome-wide significance** threshold 5 × 10⁻⁸ appear as towers because of **LD**; the SNP with the smallest p in a **locus** is the **lead SNP**. **LD clumping** reduces the hits to **independent signals**, and the results are shared as **summary statistics**. Using these SNPs as **instrumental variables** for an exposure, **Mendelian randomization** estimates its causal effect if the **three assumptions** hold, with **vertical pleiotropy** allowed but **horizontal pleiotropy** not. The **Wald ratio** gives one estimate per SNP and **IVW** combines them. In **proteomics** the exposure is **protein abundance**, the instruments are **pQTLs** (preferably **cis**, since **trans** carries more pleiotropy), and **colocalization** checks that the protein and disease signals share one causal variant. Only then does a protein move from **biomarker** to candidate **causal protein**.
:::
