---
title: "Lecture 7 — High-Dimensional Molecular Data"
subtitle: "The siNET miRNA case study · technical noise and preprocessing · moderated t (limma) · log₂FC · FWER vs FDR · Bonferroni · Benjamini–Hochberg · clustering and circular reasoning"
engine: knitr
---

```{=html}
<script>document.body.classList.add("block-omics");</script>
```

```{r}
#| include: false
library(ggplot2)
set.seed(7)
```

::: {.callout-important}
## This is the only lecture with instructor EXAM tags
The course's reading guide says: *"The slides with information that we will be testing on the exam are marked with a small text 'EXAM.'"* Every section below with the <span class="flag-badge flag-instructor">🎯 Instructor-flagged EXAM</span> badge carried that marker. **Treat these as the highest-confidence exam content in the course.** The instructor also posted **"Lecture 7 - questions.pdf"** with example exam questions. All 11 are on the [Exam practice](exam-practice.qmd) page with model answers.
:::

Up to now, each analysis tested one thing: one SNP, one exposure. Omics data flips that around. You measure hundreds or thousands of molecules (here 955 microRNAs) in a few dozen samples and ask which of them differ between groups. Three problems follow in a chain: **high dimensionality → noisy measurements → multiple testing.**

The measurements are noisy because the lab process adds its own variation, so the data must be cleaned (background correction, normalization) before any biology is compared. With few samples per feature, the ordinary t-test becomes unreliable, and the fix is to let the features borrow strength from each other (the moderated t-statistic in limma). And with 955 tests at once, about 48 will look "significant" by chance, so the significance cutoff has to change: Bonferroni controls the chance of any false positive, Benjamini–Hochberg controls the proportion of false positives among the hits.

The lecture is built around a real study: miRNA profiling of small intestinal neuroendocrine tumours (siNETs), which found tumour subtypes and a survival biomarker, miR-375.

This page follows that study from sample to result. The BH and Bonferroni calculations have worked tables, an R exercise and a drill that generates new problems. **Be able to do them by hand.**

::: {.callout-note collapse="true"}
## Sources and session
**Slides:** "Lecture 7 part 1 PRELIMINARY.pdf" (slide date 22 Sep 2026; updated the evening before the session "with some clarification and extra explanations").

**Case study article:** Arvidsson Y, Rehammar A, Bergström A, et al. *miRNA profiling of small intestinal neuroendocrine tumors defines novel molecular subtypes and identifies miR-375 as a biomarker of patient survival.* Modern Pathology 2018. The full text is not in the source guide.

**Rough plan:** Week 4, day 1. High-dimensional data (gene expression, GO, clusters), proteomics.
:::

## 7.1 Why high-dimensional data is a different problem {#t7-1 .topic flag="inferred"}

Classical epidemiology asks one question: is exposure X associated with outcome Y? Omics-scale molecular epidemiology asks: **which of hundreds or thousands of molecular features are associated with phenotype Y**, all in the same analysis.

```
High dimensionality  ──▶  Noisy measurements  ──▶  Multiple testing
   (many features,          (technical noise,         (many tests →
    few samples)             unstable variances)        many false hits)
```

Each link has its own section below. This chain is the opening framework and also the lecture's take-home message.

## 7.2 The siNET case study {#t7-2 .topic}

**Clinical background.** Small intestinal neuroendocrine tumours (siNETs) are often diagnosed only after metastasis. Five-year survival is poor but highly variable and hard to predict. Biologically they are distinctive: an unusually **low mutational burden** and **no classic recurrent driver mutations**. Tumour initiation is thought to be driven by **large-scale chromosomal alterations and epigenetic factors** instead. (A useful contrast with AML in Lecture 3: not every cancer is a point-mutation story.)

**miRNA recap** (Lecture 3): short (~22 nt) non-coding RNAs acting as epigenetic/post-transcriptional regulators: they suppress translation and reduce mRNA stability. Perfect pairing is not needed, so each miRNA regulates **hundreds of mRNAs**. Altered miRNA expression is found in virtually all tumour types. Aberrant miRNAs act as **oncomiRs** (promoting cancer) or **tumour suppressors**, by targeting genes for proliferation, cell cycle, apoptosis, invasion and drug resistance.

**Study aim:** characterize the miRNA profile of siNETs, looking for clinically relevant molecular subgroups and biomarkers of tumour development and outcome.

**Samples:** 42 tumours from 37 patients operated for siNET, plus 6 samples of normal small intestinal mucosa.

## 7.3 The microarray experiment {#t7-3 .topic}

1. **RNA isolation**: with a commercial kit (e.g. miRNeasy, Qiagen) from **fresh-frozen** tumour tissue (at least **70% tumour cells**) and normal mucosa.
2. **Labelling and hybridization**: RNA is labelled with a fluorescent dye and allowed to base-pair with known probe sequences fixed on an array (**Agilent**).
3. **Scanning**: a scanner reads fluorescence intensity at every spot. Intensity is a **relative** measure of how much of that miRNA is present; software turns the image into numbers.
4. **Design details:** **955 probes** (866 human + 89 viral miRNAs), **4–8 replicate spots** per probe (to assess consistency), **48 samples** (42 tumours + 6 normal).

## 7.4 Observed variation = biology + technical noise {#t7-4 .topic flag="instructor"}

$$\text{Observed signal} = \text{biological signal} + \text{technical variation (noise)}$$

**Sources of technical variation in this study:** RNA extraction, labelling, hybridization, scanner/detection noise and **batch effects** (developed further for proteomics in Lecture 8, Section 8.6; the principle applies to every high-throughput platform).

**Implication:** before trusting any biological comparison, remove as much technical noise as possible. That is what preprocessing does.

:::: {.quiz}
Why must microarray data be preprocessed before comparing tumour and normal samples?

- To remove the biological signal so only noise is left
- **The measured signal also contains technical variation**
- To turn relative intensities into absolute concentrations
- Because the t-test cannot be used on log-scale values

::: {.exp}
Observed variation = biology + technical noise (extraction, labelling, hybridization, scanner, batch). Preprocessing removes technical variation so remaining differences reflect biology.
:::
::::

## 7.5 Preprocessing: background correction and normalization {#t7-5 .topic flag="instructor"}

The initial analysis steps, in order:

| Step | Scope | Method in this study | What it does |
|---|---|---|---|
| 1. **Quality control** | | | |
| 2. **Background correction** | **Within** each array | **normexp** | Models observed intensity = true signal + background noise, with **noise ~ normal** and **signal ~ exponential**; estimates the expected true signal given the observed intensity. Gives **positive** corrected values (naive subtraction can give negatives) and improves reliability of **low-intensity** measurements |
| 3. **Normalization** | **Between** arrays | **Quantile normalization** | Makes the distribution of intensities **identical across arrays** |
| 4. **Average replicate probes** | Per miRNA | Mean of the 4–8 spots | One more reliable value per miRNA |

**Quantile normalization, step by step**

1. Within each sample, rank all values from lowest to highest.
2. For each rank position, compute the mean across all samples.
3. Replace each value by the mean for its rank.

It **keeps each sample's internal order** (the highest miRNA in a sample stays highest) but **equalizes the overall distribution**. Assumption: the *overall* expression distribution should be similar across arrays, so any large systematic difference in distribution shape is **technical**, not biological.

::: {.worked}
**Worked example.** Four miRNAs (A–D) on three arrays.

| miRNA | Array 1 | Array 2 | Array 3 |
|---|---|---|---|
| A | 5 | 4 | 3 |
| B | 2 | 1 | 4 |
| C | 3 | 3 | 6 |
| D | 4 | 2 | 8 |

**Sort each array and average across ranks:**

| Rank | Array 1 | Array 2 | Array 3 | Rank mean |
|---|---|---|---|---|
| 1 (lowest) | 2 | 1 | 3 | (2+1+3)/3 = **2.000** |
| 2 | 3 | 2 | 4 | (3+2+4)/3 = **3.000** |
| 3 | 4 | 3 | 6 | (4+3+6)/3 = **4.333** |
| 4 (highest) | 5 | 4 | 8 | (5+4+8)/3 = **5.667** |

**Put the rank means back in each array's original order:**

| miRNA | Array 1 | Array 2 | Array 3 |
|---|---|---|---|
| A | 5.667 (rank 4) | 5.667 (rank 4) | 2.000 (rank 1) |
| B | 2.000 (rank 1) | 2.000 (rank 1) | 3.000 (rank 2) |
| C | 3.000 (rank 2) | 4.333 (rank 3) | 4.333 (rank 3) |
| D | 4.333 (rank 3) | 3.000 (rank 2) | 5.667 (rank 4) |

Each column now contains exactly {2, 3, 4.333, 5.667}: identical distributions, original within-array order kept. Array 3's overall higher level (technical) is gone.
:::

:::: {.quiz}
After quantile normalization, which statement is true?

- Every miRNA has the same value on every array
- **Every array has the same set of values, in its own order**
- The order of miRNAs within each array is reversed
- Arrays keep their original means but get equal variances

::: {.exp}
All arrays share the same distribution (the rank means), but each array keeps its own ranking of miRNAs. A miRNA's value can still differ between arrays because its rank differs.
:::
::::

<div class="drill" data-drill="qnorm"></div>

## 7.6 Supervised vs unsupervised analysis {#t7-6 .topic}

| | Supervised | Unsupervised |
|---|---|---|
| Question | Which miRNAs differ between **predefined** groups? | Do samples **naturally** form groups based on expression? |
| Labels used? | Yes (metastasis vs primary, tumour vs normal) | No |
| Method here | Differential expression testing | Hierarchical clustering |

## 7.7 Why the ordinary t-test fails, and the moderated t {#t7-7 .topic flag="instructor"}

**The ordinary two-sample t-statistic**

$$t = \frac{\bar x_1 - \bar x_2}{s\sqrt{\tfrac{1}{n_1} + \tfrac{1}{n_2}}}$$

where $s^2$ is estimated **separately for every gene/miRNA** from only $n_1 + n_2 - 2$ degrees of freedom: just **4 df** for 3 vs 3 samples, a very typical omics size.

**Three compounding problems**

1. $s^2$ is **very noisy**: with 4 df a feature's variance estimate is often far from its true variance.
2. By chance **some features get a tiny $s^2$**. Since $s^2$ is in the denominator, $t$ becomes huge even when the mean difference is small → **false positives**.
3. The ranking is partly driven by noise in each $s^2$, so **real effects with a larger $s^2$ are missed** → lost **power**.

**The solution: the moderated t-statistic, R package `limma`.** Instead of trusting each feature's own noisy variance, limma uses **empirical Bayes** to **share information across all features**: each variance is **shrunk toward a common variance** estimated from the whole dataset. The result is a **moderated t-statistic and p-value per feature**.

The shrunken (moderated) variance is a weighted average of the common prior variance $s_0^2$ (with prior df $d_0$) and the feature's own variance $s_g^2$ (with $d_g$ residual df):

$$\tilde s_g^2 = \frac{d_0\, s_0^2 + d_g\, s_g^2}{d_0 + d_g}, \qquad \tilde t_g = \frac{\bar x_1 - \bar x_2}{\tilde s_g\sqrt{\tfrac{1}{n_1} + \tfrac{1}{n_2}}}$$

**How shrinkage behaves** (a "consequences" slide, also marked EXAM):

| Replicates | Who dominates | Effect |
|---|---|---|
| **Few** (small $n_k$) | The **global** weight $d_0$ | Each variance is pulled **strongly** toward $s_0^2$. This prevents false positives from randomly tiny variances |
| **Many** (large $n_k$) | The feature's **own** data | The moderated variance converges to the ordinary sample variance: you trust the feature's own estimate |

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 3.4
#| fig-cap: "Simulation: 2,000 features with NO true difference, 3 vs 3 samples. Left: the ordinary variance estimates scatter wildly around the true variance (dashed); moderated estimates are pulled toward the common value. Right: features with a randomly tiny s² get extreme ordinary t-values; the moderated t removes most of them."
G <- 2000; n <- 3; d <- 2 * n - 2
x1 <- matrix(rnorm(G * n, 0, 0.5), G); x2 <- matrix(rnorm(G * n, 0, 0.5), G)
diffm <- rowMeans(x1) - rowMeans(x2)
s2 <- (apply(x1, 1, var) + apply(x2, 1, var)) / 2
d0 <- 4; s02 <- median(s2)
s2t <- (d0 * s02 + d * s2) / (d0 + d)
t_ord <- diffm / sqrt(s2 * (2 / n)); t_mod <- diffm / sqrt(s2t * (2 / n))
v <- rbind(data.frame(type = "Ordinary s²", value = s2),
           data.frame(type = "Moderated s̃²", value = s2t))
v$type <- factor(v$type, levels = c("Ordinary s²", "Moderated s̃²"))
p1 <- ggplot(v, aes(value, fill = type)) +
  geom_histogram(bins = 60, alpha = 0.7, position = "identity") +
  geom_vline(xintercept = 0.25, linetype = "dashed") +
  scale_fill_manual(values = c("#8aa9c8", "#d9480f"), name = NULL) +
  labs(x = "Variance estimate", y = "Features") +
  theme_minimal(base_size = 11) + theme(legend.position = "top")
tt <- data.frame(ordinary = t_ord, moderated = t_mod, tiny = s2 < quantile(s2, 0.05))
p2 <- ggplot(tt, aes(ordinary, moderated, colour = tiny)) +
  geom_abline(colour = "grey70") +
  geom_point(size = 0.8, alpha = 0.6) +
  scale_colour_manual(values = c("grey40", "#d9480f"), labels = c("other", "5% smallest s²"), name = NULL) +
  labs(x = "Ordinary t", y = "Moderated t") +
  theme_minimal(base_size = 11) + theme(legend.position = "top")
if (requireNamespace("patchwork", quietly = TRUE)) {
  print(patchwork::wrap_plots(p1, p2))
} else {
  print(p1); print(p2)
}
```

:::: {.quiz}
In a 3 vs 3 comparison, a miRNA gets an extremely large ordinary t-value although its mean difference is small. What is the most likely reason?

- **Its variance estimate is tiny by chance, inflating t**
- Its true biological effect is larger than all others
- The array for that miRNA had too many replicate spots
- The p-value was already corrected with Bonferroni

::: {.exp}
With 4 df, some variance estimates are randomly very small. A small s² in the denominator gives a huge t: a false positive. Variance shrinkage (limma) pulls it toward the common variance.
:::
::::

:::: {.quiz}
How does limma's variance shrinkage change when the number of replicates grows large?

- Shrinkage gets stronger, pulling every variance to s₀²
- **Each feature's own variance gets more and more weight**
- The moderated t turns into a Bonferroni-corrected t
- The prior degrees of freedom d₀ grow with the sample

::: {.exp}
With many replicates the feature's own data dominate and the moderated variance converges to the ordinary sample variance. With few replicates the global prior (d₀) dominates.
:::
::::

<div class="drill" data-drill="modt"></div>

## 7.8 Summary: the small-sample problem, in full {#t7-8 .topic flag="instructor"}

Marked EXAM across three consecutive slides. **Close to a model answer for "explain why moderated statistics are needed in omics data".**

::: {.callout-note}
## Model answer structure
**1. The challenge of small sample sizes.** Transcriptomics, proteomics and metabolomics routinely measure **tens of thousands of features** on **few biological replicates**.

**2. The failure of standard t-tests.** Each feature's variance is estimated separately from very few replicates, so the estimates are **highly unstable**. By chance some are abnormally **low** → inflated t → **false positives**; others abnormally **high** → real effects masked → **lost power**. Because this instability sits under *every* p-value, it also **undermines the multiple-testing correction** afterwards, which assumes well-calibrated p-values.

**3. The standard solution.** **Sharing information across the whole dataset**, known as **variance shrinkage** or **empirical Bayes moderation**, stabilizes the variance estimates: each is pulled toward a common value estimated from all features. This **reduces false positives, recovers power** and gives **more reliable results after multiple-testing correction** (e.g. FDR).
:::

## 7.9 Log₂ transformation and log₂ fold change {#t7-9 .topic flag="instructor"}

**Log₂ transformation** makes expression values more **symmetric** and **reduces the influence of very large raw values**.

**Log₂ fold change** describes magnitude and direction of a change between groups:

::: {.worked}
| log₂FC | Fold change | Meaning |
|---|---|---|
| +3 | $2^3 = 8$ | 8× higher |
| +2 | $2^2 = 4$ | 4× higher |
| **+1** | **2** | **2× higher** |
| **0** | **1** | **No change** |
| **−1** | **1/2** | **2× lower** |
| −2 | 1/4 | 4× lower |
:::

Power-of-2 relationship: $\text{FC} = 2^{\log_2\text{FC}}$. The same logic returns in Lecture 8: one **NPX** unit (Olink/SomaScan) is a doubling.

:::: {.quiz}
A miRNA has log₂FC = −2 in metastases compared with primary tumours. What does that mean?

- It is 2 times lower in metastases
- **It is 4 times lower in metastases**
- It is 2 units lower on the raw intensity scale
- It is 4 times higher in metastases

::: {.exp}
2^−2 = 1/4, so expression in metastases is a quarter of that in primaries: 4× lower. The sign gives the direction.
:::
::::

<div class="drill" data-drill="log2fc"></div>

## 7.9b Tumour vs normal results, and the paired-sample note {#t7-9b .topic}

**Result:** **miR-7-5p** and **miR-375** were the two most significantly **upregulated** miRNAs in tumours vs normal mucosa. Several other differentially expressed miRNAs previously reported in the siNET literature also appeared (marked with an asterisk in the paper's figure).

**"Note… Why?"** Five patients had **paired** samples (a primary tumour and a liver metastasis). In the tumour vs normal comparison they contributed **only their primary tumour**.

**Why?** Two samples from the same patient are **not independent** (same patient, same genetic background). Including both would give that patient double weight and bias variance estimates and significance tests. This is a general principle: repeated or paired samples from one subject violate the independence assumption of standard tests.

## 7.10 The multiple testing problem, with this study's numbers {#t7-10 .topic flag="instructor"}

955 miRNAs were tested. If **none** truly differed, you would still expect

$$955 \times 0.05 \approx \mathbf{48}$$

"significant" hits at an uncorrected $p < 0.05$.

The logic in three steps:

1. Every test has its own **5% false-positive chance** from noise alone, even for a miRNA with zero effect.
2. Run hundreds or thousands of tests and those small chances **accumulate**: some false hits become nearly guaranteed.
3. A flat $p < 0.05$ applied test by test turns "5% error per test" into a **high chance of at least one error** somewhere in the study. For 955 independent null tests, $P(\geq 1 \text{ false positive}) = 1 - 0.95^{955} \approx 1$.

**Conclusion (stated directly in the slides): because this is multiple testing, the significance cutoff itself must change.**

## 7.11 FWER vs FDR {#t7-11 .topic flag="instructor"}

Expect to **define, distinguish and calculate** both.

::: {.worked}
| | **FWER** (Family-Wise Error Rate) | **FDR** (False Discovery Rate) |
|---|---|---|
| Definition | $P(\text{at least one false positive among all tests})$ | $E\left[\dfrac{\text{false positives}}{\text{all positives called}}\right]$, the expected **proportion** of hits that are false |
| Philosophy | Even **one** false lead among hundreds of tests is a failure of the **whole** experiment: a very strict bar | Accepts a known, controlled **fraction** of wrong hits and controls how large that fraction is expected to be |
| Classic method | **Bonferroni** | **Benjamini–Hochberg (BH)** |
:::

Both start from the **same p-values** but ask different questions, giving different, non-interchangeable cutoffs.

:::: {.quiz}
A screen calls 200 miRNAs significant with FDR controlled at 5%. What does that mean?

- The chance of any false positive among the 200 is at most 5%
- **About 10 of the 200 hits are expected to be false positives**
- Each of the 200 miRNAs has a 5% chance of being a true effect
- Exactly 5% of all 955 tested miRNAs are false positives

::: {.exp}
FDR = expected proportion of false positives among the hits: 5% of 200 ≈ 10. "Any false positive at most 5%" is FWER control (Bonferroni).
:::
::::

## 7.12 Bonferroni: controlling the FWER {#t7-12 .topic flag="instructor"}

**Rule:** reject $H_0$ for test $i$ if

$$p_i \le \frac{\alpha}{m} \qquad \text{equivalently} \qquad p_{\text{adj}} = \min(1,\; p \times m) \le \alpha$$

where $m$ is the number of tests.

**Three properties**

1. **Valid for any $m$, any test type, regardless of dependence** between tests: no assumptions about correlation, "always safe".
2. Controls the **strictest** bar, the FWER: a single false positive among all tests counts as failure.
3. **Worked consequence for this study:** $m \approx 955$ → $\alpha/m \approx 0.05/955 \approx \mathbf{0.00005}$. Many true effects do not survive this, so **power collapses** when $m$ is large, which is typical for omics.

## 7.13 Benjamini–Hochberg: controlling the FDR {#t7-13 .topic flag="instructor"}

**The BH algorithm, three steps (reproduce exactly):**

1. Sort the $m$ p-values: $p_{(1)} \le p_{(2)} \le \dots \le p_{(m)}$.
2. Find the **largest rank $k$** with $p_{(k)} \le \dfrac{k}{m}\,Q$, where $Q$ is the accepted FDR (commonly 0.05).
3. Call **every test with $p \le p_{(k)}$** significant. The expected FDR of this set is $\le Q$.

**Key difference from Bonferroni:** Bonferroni has one **fixed** threshold $\alpha/m$. BH's threshold $\frac{k}{m}Q$ **rises with rank**, so larger p-values can still be significant if they clear the higher bar for their rank. That is why BH finds more true positives.

::: {.worked}
**🎯 Instructor-flagged worked example: m = 8, Q = 0.05**

| Rank $i$ | $p_{(i)}$ | Threshold $\frac{i}{m}Q$ | $p_{(i)} \le$ threshold? | Significant? |
|---|---|---|---|---|
| 1 | 0.001 | 0.00625 | yes | ✓ |
| 2 | 0.011 | 0.01250 | yes | ✓ |
| 3 | 0.015 | 0.01875 | yes | ✓ |
| 4 | 0.024 | 0.02500 | yes | ✓ |
| 5 | 0.041 | 0.03125 | no | — |
| 6 | 0.045 | 0.03750 | no | — |
| 7 | 0.049 | 0.04375 | no | — |
| 8 | 0.060 | 0.05000 | no | — |

Largest $k$ with $p_{(k)} \le$ threshold: **$k = 4$** → tests 1–4 significant.

**Bonferroni on the same data:** $\alpha/m = 0.05/8 = 0.00625$ → **only test 1** significant.
:::

::: {.callout-warning}
## The "largest k" trap
BH does not stop at the first p-value that fails. You look for the **largest** rank that passes. If rank 5 failed but rank 6 passed, then **all of ranks 1–6** would be significant, including rank 5. (In the example above no rank after 4 passes, so k = 4.)
:::

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 3.6
#| fig-cap: "The same 8 p-values against the two rules. BH's threshold (green) rises with rank; Bonferroni's (red) is flat at α/m."
bh <- data.frame(rank = 1:8, p = c(0.001, 0.011, 0.015, 0.024, 0.041, 0.045, 0.049, 0.060))
bh$sig <- ifelse(bh$rank <= 4, "BH significant", "not significant")
ggplot(bh, aes(rank, p)) +
  geom_line(aes(y = rank / 8 * 0.05, linetype = "BH: (i/m)·Q"), colour = "#2f9e44", linewidth = 1) +
  geom_hline(aes(yintercept = 0.05 / 8, linetype = "Bonferroni: α/m"), colour = "#d9480f", linewidth = 1) +
  geom_point(aes(colour = sig), size = 3.5) +
  scale_colour_manual(values = c("#2f9e44", "grey50"), name = NULL) +
  scale_linetype_manual(values = c("solid", "dashed"), name = NULL) +
  scale_x_continuous(breaks = 1:8) +
  labs(x = "Rank i", y = "p-value") +
  theme_minimal(base_size = 12) + theme(legend.position = "top", legend.box = "vertical")
```

:::: {.quiz}
Five tests give the p-values 0.030, 0.004, 0.300, 0.026, 0.045. How many are significant with BH at Q = 0.05?

- 1
- 2
- **3**
- 4

::: {.exp}
Sort: 0.004, 0.026, 0.030, 0.045, 0.300. Thresholds (i/5)·0.05: 0.01, 0.02, 0.03, 0.04, 0.05. Rank 1 passes, rank 2 fails (0.026 > 0.02), rank 3 passes (0.030 ≤ 0.03), ranks 4 and 5 fail. The largest passing rank is k = 3, so the three smallest p-values are significant, including 0.026, which failed its own threshold. Stopping at the first failure (answer 1) is the classic mistake.
:::
::::

## 7.13b BH-adjusted p-values {#t7-13b .topic flag="instructor"}

BH can also be written as an **adjusted p-value per test**:

$$p^{\text{adj}}_{(i)} = \min_{j \ge i}\left( p_{(j)} \cdot \frac{m}{j} \right)$$

Take the **smallest** value of $p_{(j)} \cdot m/j$ over all ranks from $i$ **down to the bottom** of the list. This "running minimum from the bottom" makes the adjusted p-values **non-decreasing** down the sorted list, which $p_{(i)} \cdot m/i$ alone does not guarantee.

::: {.worked}
**🎯 Instructor-flagged worked example, same 8 tests**

| Rank $i$ | $p_{(i)}$ | $p_{(i)} \cdot m/i$ | Adjusted p | Significant ($\le 0.05$)? |
|---|---|---|---|---|
| 1 | 0.001 | 0.008 | **0.008** | ✓ |
| 2 | 0.011 | 0.044 | **0.040** | ✓ |
| 3 | 0.015 | 0.040 | **0.040** | ✓ |
| 4 | 0.024 | 0.048 | **0.048** | ✓ |
| 5 | 0.041 | 0.066 | **0.056** | — |
| 6 | 0.045 | 0.060 | **0.056** | — |
| 7 | 0.049 | 0.056 | **0.056** | — |
| 8 | 0.060 | 0.060 | **0.060** | — |

How to fill the "Adjusted p" column: start at the **bottom** (rank 8: 0.060). Move up; each value is the minimum of its own $p \cdot m/i$ and the adjusted value below it. Rank 7: min(0.056, 0.060) = 0.056. Rank 6: min(0.060, 0.056) = 0.056. Rank 5: min(0.066, 0.056) = 0.056. Rank 4: min(0.048, 0.056) = 0.048. Rank 3: min(0.040, 0.048) = 0.040. Rank 2: min(0.044, 0.040) = **0.040**. Rank 1: min(0.008, 0.040) = 0.008.
:::

Same decision as the threshold version (tests 1–4), but now each test has its own adjusted p-value: the format reported in most papers, including the siNET paper. In R: `p.adjust(p, method = "BH")`.

## 7.14 Why FDR fits omics data {#t7-14 .topic flag="instructor"}

**Four reasons, a strong essay-question candidate ("why is FDR preferred over Bonferroni in omics?"):**

1. **Too many tests for Bonferroni.** Thousands of tests push α/m toward zero, so most true effects never reach it.
2. **Exploratory, not confirmatory.** The goal is a **shortlist of candidates**, not one make-or-break claim. A small, controlled fraction of false leads is an acceptable price for finding more true positives.
3. **Findings are validated afterwards** with an independent method and/or cohort. The siNET study did exactly this: **miR-375 was confirmed with qRT-PCR and in situ hybridization in a separate cohort.**
4. **Features are correlated.** Co-regulated miRNAs (genes, proteins) mean the *effective* number of independent tests is smaller than $m$. Bonferroni ignores this and stays needlessly strict; BH copes better in practice.

:::: {.quiz}
Which is NOT one of the four reasons FDR fits omics data?

- Thousands of tests make Bonferroni's threshold too strict
- Screens are exploratory, and hits are validated afterwards
- Correlated features make the effective number of tests smaller
- **FDR guarantees that no false positives are among the hits**

::: {.exp}
FDR accepts a controlled proportion of false positives; it never guarantees zero. Even FWER control only bounds the probability of any false positive.
:::
::::

## 7.15 Multiple testing in the siNET paper {#t7-15 .topic}

Quote from the methods: **"All P-values were adjusted for multiple testing using Benjamini–Hochberg false discovery rate."**

| Tumour vs normal | |
|---|---|
| miRNAs tested | 955 |
| Significant (adjusted P < 0.05) | **155** (75 up, 80 down) |
| miR-375 adjusted P | **4.2 × 10⁻¹⁹** |

## 7.16 Cluster analysis and downstream findings {#t7-16 .topic}

**Unsupervised hierarchical clustering** of miRNA profiles with **complete linkage** and **Euclidean distance**. Associations between clusters and clinical variables were tested with **Fisher's exact test** and the **t-test**. Samples separated into normal (N), primary (P) and metastatic (M) groups; metastases split further into **M1 and M2**.

**M1 vs M2**

| Feature over-represented in M1 | P |
|---|---|
| Higher grade (G2/G3), i.e. higher proliferation | 0.0011 |
| Ki67 index > 3% (proliferation) | 0.00023 |
| Gain of chromosome 14 | 0.0013 |

(Chromosome 14 gain ties back to siNETs being driven by chromosomal alterations, Section 7.2.)

**Differential expression M1 vs M2:** 76 miRNAs up and 65 down in M1 (adjusted P < 0.05). Most upregulated in M1: **miR-1246** and **miR-663a**, highly expressed in several cancers and implicated as **oncomiRs** (proliferation, invasiveness, metastasis). Most downregulated: **miR-488-3p**, reported as a **tumour suppressor** in gastric cancer.

**Survival by cluster:** patients with **M1** tumours had **shorter overall survival** than M2 (log-rank P = 0.029; from the instructor's example questions).

**Survival analysis per miRNA:** **one Cox regression per miRNA**, adjusted for age and gender. **miR-375** was the top candidate: **higher expression → longer survival**. miR-375 was also shown to be expressed in liver metastases via tissue microarray.

## 7.16b Circular reasoning: testing clusters with the data that built them {#t7-16b .topic flag="instructor"}

The M1/M2 clusters were **created from miRNA expression** (unsupervised clustering), and then miRNA expression was **tested again** for differences **between those same clusters**. The same data is used twice: once to define the groups, again to test them.

**This is circular reasoning and inflates significance.** Any clustering produces groups that differ in the variables used to build it, almost by construction, so "finding" that they differ is not independent evidence.

::: {.callout-important}
## General principle
Do not use one dataset to both **discover** a grouping and **test** for differences across that grouping without independent validation (a new cohort) or a correction for the selection.
:::

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 3.3
#| fig-cap: "Pure noise: 500 'miRNAs' in 30 samples with no real groups. Cluster the samples into two groups, then t-test every miRNA between those clusters. Under the null the p-values should be flat (dashed line); instead they pile up near zero."
X <- matrix(rnorm(30 * 500), nrow = 30)
cl <- cutree(hclust(dist(X), method = "complete"), k = 2)
if (min(table(cl)) < 3) cl <- kmeans(X, 2)$cluster
pv <- apply(X, 2, function(g) t.test(g[cl == 1], g[cl == 2])$p.value)
ggplot(data.frame(p = pv), aes(p)) +
  geom_histogram(breaks = seq(0, 1, 0.05), fill = "#d9480f", colour = "white") +
  geom_hline(yintercept = 500 / 20, linetype = "dashed") +
  labs(x = "p-value (cluster 1 vs cluster 2)", y = "miRNAs",
       subtitle = paste0(sum(pv < 0.05), " of 500 'significant' at p < 0.05 (expected by chance: 25)")) +
  theme_minimal(base_size = 12)
```

:::: {.quiz}
Why are the M1 vs M2 differential expression results over-optimistic?

- The Cox regression was not adjusted for age and gender
- Hierarchical clustering cannot use Euclidean distance
- **M1 and M2 were defined from the same miRNA data**
- Benjamini–Hochberg is too liberal for 955 miRNAs

::: {.exp}
The clusters were built from the expression data, so they differ in that data almost by construction. Testing the same data between those clusters is circular and inflates significance.
:::
::::

## 7.17 Take-home message {#t7-17 .topic}

The lecture's own closing slide:

1. **High dimensionality → noisy measurements → multiple testing** (7.1).
2. **Observed variation = biological + technical variation**, requiring preprocessing; quantile normalization as the worked example (7.4–7.5).
3. **The ordinary t-statistic fails at small sample sizes; the moderated t-statistic is the solution** (7.7).
4. **Multiple testing: FWER vs FDR**, and being able to **calculate Bonferroni- and BH-adjusted values** (7.10–7.13).
5. **Validation in an independent cohort with an independent method** before treating a screening hit as established (7.14).

::: {.callout-note collapse="true"}
## Consolidated list of all 🎯 EXAM topics in Lecture 7
- Observed variation = biology + technical noise
- Preprocessing: QC → background correction (normexp) → normalization (quantile) → averaging replicates
- Why the ordinary t-test fails at small sample sizes
- Moderated t / limma / empirical Bayes shrinkage; behaviour at small vs large sample sizes
- The three-part summary: challenge of small samples / failure of standard t-tests / standard solution
- Log₂ transformation and log₂ fold change
- The multiple testing problem (955 × 0.05 ≈ 48)
- FWER vs FDR
- Bonferroni and its consequence (0.05/955 ≈ 0.00005)
- BH, threshold version and adjusted p-values, with worked tables
- Four reasons FDR fits omics
- Circular reasoning when testing clusters built from the same data
:::

# Practice

::: {.callout-note}
## Benjamini–Hochberg and Bonferroni: generated problems
Each click on **New problem** draws new p-values. Do it on paper first, then type your answers and press **Check**. The solution shows the full sorted table with thresholds and adjusted p-values.
:::

<div class="drill" data-drill="bh"></div>

<div class="drill" data-drill="bh"></div>

::: {.callout-note}
## BH, Bonferroni and the moderated t in R
Runs in your browser (the first run takes a few seconds while R loads).
:::

```{webr-r}
#| label: bh-practice
p <- c(0.001, 0.011, 0.015, 0.024, 0.041, 0.045, 0.049, 0.060)
m <- length(p); Q <- 0.05

# --- BH by hand ---------------------------------------------------------
ps <- sort(p)
rank <- seq_len(m)
threshold <- rank / m * Q
passes <- ps <= threshold
k <- if (any(passes)) max(rank[passes]) else 0   # LARGEST passing rank
raw_adj <- ps * m / rank
adj <- rev(cummin(rev(raw_adj)))                  # running min from the bottom
data.frame(rank, p = ps, threshold, passes, raw_adj = round(raw_adj, 4),
           adj = round(adj, 4), significant = rank <= k)

# --- Compare with R's built-in functions --------------------------------
round(p.adjust(p, method = "BH"), 4)
round(p.adjust(p, method = "bonferroni"), 4)
sum(p.adjust(p, method = "BH") <= 0.05)
sum(p <= 0.05 / m)
```

**Tasks**

1. Change the 5th p-value to `0.030`. What is $k$ now, and which tests are significant?
2. Now set `p <- c(0.002, 0.020, 0.013, 0.030, 0.200)` (unsorted on purpose). Work out BH on paper first, then check.
3. With 955 tests, what is the Bonferroni threshold? Compute `0.05 / 955`.
4. Simulate 955 null tests: `p <- runif(955)`. How many have $p < 0.05$? How many survive BH and Bonferroni?

::: {.callout-tip collapse="true"}
## Model answer
1. Sorted: 0.001, 0.011, 0.015, 0.024, **0.030**, 0.045, 0.049, 0.060. Thresholds: 0.00625, 0.0125, 0.01875, 0.025, **0.03125**, 0.0375, 0.04375, 0.05. Rank 5 now passes (0.030 ≤ 0.03125), later ranks fail → **k = 5**, tests 1–5 significant.
2. Sorted: 0.002, 0.013, 0.020, 0.030, 0.200; thresholds 0.01, 0.02, 0.03, 0.04, 0.05. Ranks 1–4 all pass, rank 5 fails → **k = 4**. Bonferroni (0.01): only 0.002.
3. 0.05/955 ≈ **5.2 × 10⁻⁵** (≈ 0.00005).
4. About 48 have p < 0.05 (955 × 0.05); typically **0** survive BH or Bonferroni, as they should when nothing is real.
:::

```{webr-r}
#| label: modt-practice
# Ordinary vs moderated t on pure noise: 3 vs 3 samples, 2000 features
set.seed(1)
G <- 2000; n <- 3; d <- 2 * n - 2
x1 <- matrix(rnorm(G * n), G); x2 <- matrix(rnorm(G * n), G)
diff <- rowMeans(x1) - rowMeans(x2)
s2   <- (apply(x1, 1, var) + apply(x2, 1, var)) / 2   # pooled per-feature variance

d0  <- 4                 # prior degrees of freedom
s02 <- mean(s2)          # common (prior) variance
s2_mod <- (d0 * s02 + d * s2) / (d0 + d)

t_ord <- diff / sqrt(s2 * 2 / n)
t_mod <- diff / sqrt(s2_mod * 2 / n)
p_ord <- 2 * pt(-abs(t_ord), df = d)
p_mod <- 2 * pt(-abs(t_mod), df = d + d0)

c(ordinary_hits = sum(p_ord < 0.001), moderated_hits = sum(p_mod < 0.001))
head(cbind(s2, s2_mod, t_ord, t_mod)[order(s2), ], 5)   # the smallest variances
```

**Tasks**

5. Look at the five features with the smallest $s^2$. What happens to their t-values after moderation?
6. Increase `n` to 20 and rerun. Does moderation still change much? Why?

::: {.callout-tip collapse="true"}
## Model answer
5. Their tiny variances are pulled up toward $s_0^2$, so their inflated ordinary t-values shrink to ordinary sizes. Moderation removes these chance "hits".
6. With `n = 20`, d = 38 dominates d₀ = 4, so $\tilde s^2 \approx s^2$ and the two t-statistics nearly agree. Shrinkage matters most when samples are few.
:::

# Why-questions

::: {.callout-tip collapse="true"}
## Why does BH find more hits than Bonferroni on the same data?
Bonferroni applies the same strict threshold α/m to every test. BH's threshold (i/m)·Q grows with rank, and it controls a weaker error rate (the expected proportion of false hits, not the chance of any false hit).
:::

::: {.callout-tip collapse="true"}
## Why is quantile normalization justified?
It assumes that the overall distribution of expression should be about the same in every sample, so systematic differences in distribution are technical. If that assumption fails (e.g. a global biological shift in expression), quantile normalization would remove real biology.
:::

::: {.callout-tip collapse="true"}
## Why is the moderated t more powerful, not just more conservative?
Shrinkage also pulls randomly *large* variances down toward the common value, so real effects that were masked by an unlucky high s² can now be detected. And the moderated t has more degrees of freedom (d + d₀).
:::

::: {.callout-tip collapse="true"}
## Why was only the primary tumour used from paired patients?
A primary tumour and its metastasis come from the same patient and are not independent. Using both would double-count that patient and bias the variance estimates and tests.
:::

::: {.callout-tip collapse="true"}
## Why is a significant difference between clusters not a finding?
The clusters were built to be different in exactly those miRNAs. The difference is guaranteed by the method, not discovered in the data. Pure noise gives the same "result" (see the histogram in 7.16b).
:::

# Spaced review

:::: {.quiz}
From Lecture 4: the genome-wide significance threshold 5 × 10⁻⁸ is based on which correction?

- Benjamini–Hochberg at 5% FDR
- **Bonferroni for about 1 million tests**
- A moderated t-test with shrinkage
- Quantile normalization of p-values

::: {.exp}
0.05 / 1,000,000 = 5 × 10⁻⁸: a Bonferroni-style correction for about one million independent SNPs.
:::
::::

:::: {.quiz}
From Lecture 3: how do miRNAs differ from siRNAs?

- miRNAs match their target perfectly and silence one gene
- **miRNAs bind imperfectly and regulate many mRNAs**
- miRNAs come from double-stranded RNA made in the lab
- miRNAs are translated into small regulatory proteins

::: {.exp}
miRNA binds imperfectly, so one miRNA regulates many targets; siRNA binds near-perfectly and silences one.
:::
::::

# Further reading {#further-reading}

The course's reading guide summarizes the lecture's main themes (last slide, page 51): noisy data due to technical variation, which needs quality control and preprocessing; high dimensionality with few samples, where the moderated t-statistic replaces the ordinary t; and multiple testing, FWER and FDR, Bonferroni and Benjamini–Hochberg. *"The slides with information that we will be testing on the exam are marked with a small text 'EXAM'."*

- **Arvidsson Y, Rehammar A, Bergström A, et al.** miRNA profiling of small intestinal neuroendocrine tumors defines novel molecular subtypes and identifies miR-375 as a biomarker of patient survival. *Modern Pathology* 2018. The article the case study is built on (PDF on Canvas below).
- **"Lecture 7 - questions.pdf"**: all 11 questions are worked through on the [Exam practice](exam-practice.qmd) page.

::: {.callout-note collapse="true"}
## Course files on Canvas (GU login required)
- [Lecture 7.pdf (updated 22 Sep, 16.30)](https://canvas.gu.se/courses/101301/files/11714353?wrap=1){target="_blank"}
- [miRNA_profiling_of_siNETs_defines_molecular_subtypes_and_identifies_biomarker_of_patient_survival.pdf](https://canvas.gu.se/courses/101301/files/11709324?wrap=1){target="_blank"}
- [Lecture 7 - questions.pdf](https://canvas.gu.se/courses/101301/files/11714701?wrap=1){target="_blank"}
:::
